Sources of Measurement Bias in Modern CPUs

### 2014-02-02 paper.tex
Add analysis stuff on loop

\lstset{language=C, basicstyle=\ttfamily, columns=flexible}

@@

\subsection{Bias from Environment Size}
Variations in performance with bias towards certain environment sizes has been observed by other authors.
Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack. 
Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at run time, assuming a 64 bit process mapped to virtual memory.

@@

\subsection{Analysis}

A small amount of assembly code was used to calculate the addresses of each stack allocated variable and output them directly using system calls, without impacting run time addresses.
We find that the spike in cycle count occurs when the address of g and inc alias with i and j, respectively.
Resource stalls are generated because of false dependencies between the addresses of (g, i) and (inc, j), resulting in worse performance.
Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment.
The program is \emph{biased} towards environment sizes that avoids this specific stack alignment.

@@

\begin{lstlisting}[float=h, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopkernel}, frame=lines]
static int i, j, k;

int main() {
    int g = 0, inc = 1;

    if ((((long)&inc) & 0xfff) == (((long)&i) & 0xfff) || 
        (((long)&g) & 0xfff) == (((long)&i) & 0xfff))
        return main();

    for (; g < 65536*10; g++) {
        i += inc;
        j += inc;
        k += inc;
    }

    return 0;
}
\end{lstlisting}

\subsection{Heap Address Aliasing}
Address aliasing can be caused by conflicting pairs of load/store operations to any part of memory.
The previous section looked specifically at an example of collision between stack variables and static data.
Another scenario to consider is collisions in heap allocated memory.
In particular, code that operate on pairs of contiguous arrays can be vulnerable to 4K aliasing.
Acquiring dynamic memory at run time is usually done by calling malloc, which takes a number of bytes to allocate and returns a pointer to that area.
A typical malloc implementation uses two different strategies for how to allocate memory.
The \emph{heap} area in virtual memory is used for smaller allocations.
Programs are initially given a small heap area, which the allocator can increase by calling brk or sbrk.
For larger allocations, it is typically faster to use anonymous memory mapping by calling mmap, instead of growing the ``normal'' heap.
The heap starts at a low address close to the data section and grows towards higher addresses. 
Mmap allocates much closer to the stack.
While heap addresses can look like 0x16e30a0 or 0x1723020, pointers returned by mmap can for example be 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, mmap has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by mmap will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of medatada at the beginning, therefore every mmap'ed address ends with 0x010}

@@

>>>>

\lstset{language=C, basicstyle=\ttfamily\small, columns=flexible}

@@

\section{Bias from Environment Size}
Variations in performance with bias towards certain environment sizes has been observed by other authors.
Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack. 
Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at run time, assuming a 64 bit process mapped to virtual memory.

@@

\subsection{Analysis}
To be able to explain what is happening in this case we need to know the addresses of all variables at run time.
For statically allocated data, memory allocation is decided at compile time.
The location of variables i, j and k can be found by looking at the symbol table in the ELF executable.
Below is an excerpt of \lstinline$readelf -s$.
\begin{lstlisting}[float=h, basicstyle=\small]
37: 000000000060103c    4  (...)  25 i
38: 0000000000601040    4  (...)  25 j
39: 0000000000601044    4  (...)  25 k
\end{lstlisting}

A small amount of assembly code was used to calculate the addresses of each stack allocated variable and output them directly using system calls, without impacting run time addresses.
We find that the spike in cycle count occurs when the address of inc alias with i, where the first spike happens when inc is at 0x7fffffffe03c.
Because the stack is aligned to 16 byte or 4 words, there are a couple of different scenarios that could have been observed here.
Static area is fixed and covers 12 contiguous bytes (3 words) \lstinline$[x|x| |x]$.
Automatic variables occupies 8 contiguous bytes on stack (2 words) [ | |x|x]. 
The space of possible addresses for (inc, g) in this case is (0x7fffffffe03c, 0x7fffffffe038) - n*4096. 
In this scenario, g will never alias with any of the static variables -- as it always covers the 0x8 slot not occupied by either of i, j or k.
An even worse case with respect to the number of alias events is if allocation of static data allows collisions with both stack allocated variables.
This can be achieved by reserving 8 extra bytes to offset i, j into the 0x8, 0xc slots. 
While this will give significantly more alias counts, it has little effect on the number of cycles executed.

Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment.
Resource stalls are generated because of false dependencies between stack and static data, resulting in worse performance.
The program is \emph{biased} towards environment sizes that avoids this specific stack alignment.

@@

\begin{lstlisting}[float=h, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopkernel}, frame=lines]
#define ALIAS(a, b) \
    ((((long)&a)&0xfff)==(((long)&b)&0xfff))
static int i, j, k;
int main() {
    int g = 0, inc = 1;
    if (ALIAS(inc, i) || ALIAS(g, i))
        return main();
    for (; g < 65536*10; g++) {
        i += inc;
        j += inc;
        k += inc;
    }
    return 0;
}
\end{lstlisting}

\section{Bias from Heap Allocations}
Address aliasing can be caused by conflicting pairs of load/store operations to any part of memory.
The previous section looked specifically at an example of collision between stack variables and static data.
Another scenario to consider is collisions in heap allocated memory.
In particular, code that operate on pairs of contiguous arrays can be vulnerable to 4K aliasing.
Acquiring dynamic memory at run time is usually done by calling malloc, which takes a number of bytes to allocate and returns a pointer to that area.
A typical malloc implementation uses two different strategies for how to allocate memory.
The \emph{heap} area in virtual memory is used for smaller allocations.
Programs are initially given a small heap area, which the allocator can increase by calling brk or sbrk.
For larger allocations, it is typically faster to use anonymous memory mapping by calling mmap, instead of growing the ``normal'' heap.
The heap starts at a low address close to the data section and grows towards higher addresses. 
Mmap allocates much closer to the stack.
While heap addresses can look like 0x16e30a0 or 0x1723020, pointers returned by mmap can for example be 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, mmap has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by mmap will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of medatada at the beginning, therefore every mmap'ed address ends with 0x010}

@@

\section{Aliasing in BLAS}
Another test case, show numbers for ATLAS OR default blas in Ubuntu, or maybe both. Copy thesis more or less I guess.
